An accelerated algorithm for finding efficient solutions in multiobjective problems with black-box multiextremal criteria

### 2022-01-14 2022/OM&S/paper.tex
almost done

\title{A fast algorithm for finding efficient solutions in multi-objective problems with black-box multiextremal criteria}

\author{
\name{Konstantin Barkalov, Vladimir Grishagin and Evgeny Kozinov\thanks{CONTACT Konstantin Barkalov. Email: konstantin.barkalov@itmm.unn.ru}}
\affil{Software Department, Lobachevsky State University, Nizhni Novgorod, Russia}
}

@@

\begin{abstract}
The paper is devoted to consideration of multiextremal black-box MCO problems with time-consuming criteria and proposes a new computational algorithm for solving such problems. This algorithm is based on combination of convolution approach, dimensionality reduction, and information-statistical global optimization algorithm. For approximating the Pareto set a family of maximum convolutions is considered generating problems of scalar multiextremal optimization. The latters are reduced by means of Peano mappings to the equivalent problems of univariate optimization that are solved by a modification of the information-statistical global optimization method. This method applies the technique of local refinements and uses in the course of current univariate optimization the information about criteria that was obtained when solving completed optimization problems. These features allow one to accelerate the search of global solution.  
The efficiency of the proposed MCO algorithm is experimentally estimated on representative classes of multidimensional multiextremal test problems. Comparison with 5 nature-inspired particle swarm and genetic algorithms on the base of the hypervolume index reflecting quality of Pareto set evaluation demonstrates qualitative results of the proposed method.
\end{abstract}

@@

In the field of mathematical modeling the intelligent decision-making processes, the problems of multicriterial optimization (MCO) present a wide class of models that are rich and interesting from a theoretical point of view and very important in real applications \cite{Marler2009,Hillermeier2005}. Multicriterial problems describe decision-making situations with multiple goals which are considered as functional criteria to be optimized.

@@

Finally, the last factor is the multiextremality of criteria \cite{Pardalos2017}. First of all, multiextremal criteria can generate complicated Pareto sets consisting, for example, of several disconnected parts. Moreover, the combination of such factors as multiextremality and greater dimension leads to significant computational complexity of the problem because the computational cost increases exponentially when dimension grows. From this point of view the multiextremal problems are the most complicated ones in multicriterial optimization and this MCO class is the subject of interest in the given paper.

@@

Another idea applies a parameterization scheme for the criteria of the problem and reduces a MCO problem to a family of single-criterion (scalar) optimization problems being problems of mathematical programming which can be solved by corresponding well-developed optimization methods \cite{Collette2004,Ehrgott2005}. In this scheme a non-negative coefficient (weight) is assigned to each criterion which reflects numerically an ``importance'' of the criterion and on the base of such weighted criteria a scalar function (convolution) is built and minimized. Under appropriate choice of convolution the solution of the built scalar optimization problem is a partial solution of the initial multicriterial problem. Changing weights we can obtain different Pareto points.

@@

The problem of multicriterial optimization is formulated as minimizing over the domain (\ref{Q}) the vector function
\begin{equation}\label{W}
W(x) = \left(w_1(x), ..., w_k(x)\right).
\end{equation}

@@

As mentioned in Introduction, one way of building the Pareto set consists in reducing the multiobjective problem to a family of single-criteria, or scalar optimization problems and solving them by methods of mathematical programming.  The global minimum point of a problem from this family will be a Pareto optimal point under corresponding assumptions. This approach can be realized by means of applying a convolution of the vector criterion (\ref{W}), for example, the maximum convolution
\begin{equation}\label{conv}
\Gamma(\lambda,x) = \max_{1 \leq i \leq k}{\lambda_i w_i(x)},
\end{equation}
where coefficients $\lambda_i, \; 1 \leq i \leq k,$ (weights of criteria) satisfy the conditions
\begin{equation}\label{lambda}
\lambda_i \geq 0, \; 1 \leq i \leq k, \; \sum_{i=1}^k{\lambda_i} = 1.
\end{equation}

If
\begin{equation}\label{positive}
w_i(x) > 0, \; 1 \leq i \leq k,
\end{equation}
in the domain (\ref{Q}), then global solution to the problem
\begin{equation}\label{conv_problem}
\Phi_\lambda(x) = \Gamma(\lambda,x) + \gamma \sum_{i=1}^k{\lambda_i w_i(x)} \rightarrow \min, \; x \in Q,
\end{equation}
with a sufficiently small parameter $\gamma > 0$ is a Pareto optimal point of the initial MCO problem (\ref{problem}) \cite{Wierzbicki} [Wierzbicki, Krasnoshekov]. The positiveness of criteria is not restrictive requirement because it is possible to transform easily the MCO problem with non-positive criteria to the form (\ref{positive}) without loss of Pareto solution.

@@

If the function $f(x)$ satisfies the Lipschitz condition (\ref{lip_1}) then $H=2L\sqrt{N+3}$.

As a consequence, minimization of the function $f(x(t))$  requires the use of special algorithms oriented at functions with property (\ref{holder}).

In the proposed scheme for solving reduced univariate minimization problem we apply modified information-statistical algorithm \cite{Strongin2000} with local refinements (ALR) accelerating convergence to global optimum of functions that satisfy the H{\"o}lder condition.

@@

\item
The new $(n+1)$-th trial is performed at the point
\begin{equation}\label{eq:22}
t^{n+1} = \frac{t_p-t_{p-1}}{2} - sign(v_p-v_{p-1}) \left( \frac{|v_p-v_{p-1}|}{M} \right)^N \frac{1}{2r} 
\end{equation}
the value $v^{n+1} = \varphi(t^{n+1})$ is computed and the iteration number $n$ is increased by 1.

@@

The method ALR described above is a modification of the basic information-statistical algorithm \cite{Strongin2000,Sergeyev2013} for minimization functions satisfying the H{\"o}lder condition. The modification consists in introducing new characteristics (\ref{eq:17}) after initial stage of the search to improve the method's speedup on account of increasing values of characteristics for intervals containing current record (\ref{eq:18}). 
\textcolor[rgb]{1,0,0}{Actually, the second term in (\ref{eq:17}) will have the largest value equal to $(1.5)^{\sigma}$ only for those intervals for which the equality $v_q = \varphi^*(n)$ or $v_{q-1} = \varphi^*(n)$ holds, i.e. for the intervals with the boundary points correspond to the current record $\varphi^*(n)$.}

@@

\textcolor[rgb]{1,0,0}{The complete convergence theory of the information-statistical algorithm is given in \cite{Strongin2000,Sergeyev2013}.}

@@

During all experiments the method ALR was used with parameters $r=5.6$, $\sigma = 18$, $T=3$. 
\textcolor[rgb]{1,0,0}{The use of $T=3$ corresponds to alternation of one ``local'' iteration and two ``global'' iterations, that can be interpreted as a combination of local refinement and global updating of the current optimizer.}
Accuracy $\varepsilon$ from the stopping condition (\ref{eq:23}) was $0.05$ in bi-criteria problems and $0.01$ in 4-criteria ones.

@@

\textcolor[rgb]{1,0,0}{The Pareto front in one of test problems is shown in Fig.~\ref{fig2a} and Fig.~\ref{fig2b}. Here Fig.~\ref{fig2a} corresponds to the Pareto front computed on the uniform grid with the step 0.05 in the search domain whereas Fig.~\ref{fig2b} shows the Pareto front computed by RALR-DA. In the last case it is required only 608 trials to obtain the solution with the same accuracy.}

@@

\textcolor[rgb]{1,0,0}{The averaged HV indices obtained while solving the two-dimensional bi-criteria problems are presented} in Tab.~\ref{tab1} and Fig.~\ref{fig3} where ``Pareto'' corresponds to the estimation obtained by the grid method.

@@

\begin{table}
\tbl{Results of testing bi-criteria 4-dimensional MCO problems.}
{\begin{tabular}{ccccccc} \toprule
 Pareto & RALR-DA & OMOPSO & SPEA2 & SMPSO & IBEA & NSGAII   \\ \midrule
  23.89 & 23.52 & 22.87 & 22.80 & 22.53 & 20.68 & 19.82 \\ \bottomrule
\end{tabular}}
\label{tab3}
\end{table}

@@

As the directions of further development of the current research it would be interesting to investigate the possibilities of applying other schemes of dimensionality reduction, for example,
\textcolor[rgb]{1,0,0}{the nested optimization scheme \cite{Grishagin2016,Grishagin2018}, the diagonal \cite{Sergeyev2015,Sergeyev2017} or the simplicial \cite{Zilinskas2011,PaulaviciusZilinskas2014} partitions methods}
in the framework of the general approach used in the paper. Moreover, the prospective way is to develop parallel versions of the proposed general scheme oriented at powerful supercomputer systems.

>>>>

\title{An accelerated algorithm for finding efficient solutions in multiobjective problems with black-box multiextremal criteria}

\author{
\name{Konstantin Barkalov, Vladimir Grishagin and Evgeny Kozinov\thanks{CONTACT Konstantin Barkalov. Email: konstantin.barkalov@itmm.unn.ru}}
\affil{Institute of Information Technology, Mathematics and Mechanics, Lobachevsky State University, 23 Gagarin ave., 603022 Nizhny Novgorod, Russia}
}

@@

\begin{abstract}
The paper is devoted to consideration of multiextremal multicriteria black-box problems with time-consuming criteria and proposes a new computational algorithm for solving such problems. This algorithm is based on combination of convolution approach, dimensionality reduction, and information-statistical global optimization algorithm. For approximating the Pareto set a family of maximum convolutions is considered generating problems of scalar multiextremal optimization. The latters are reduced by means of Peano mappings to the equivalent problems of univariate optimization that are solved by an accelerated information-statistical global optimization method. This method applies the technique of local refinements and uses in the course of current univariate optimization the information about criteria that was obtained when solving completed optimization problems. These features allow one to accelerate the search of global solution.  
The efficiency of the proposed algorithm is experimentally estimated on representative classes of multidimensional multiextremal test problems. Comparison with 5 nature-inspired particle swarm and genetic algorithms on the base of the hypervolume index reflecting quality of Pareto set evaluation demonstrates qualitative results of the proposed method.
\end{abstract}

@@

In the field of mathematical modeling the intelligent decision-making processes, the problems of multicriteria optimization (MCO) present a wide class of models that are rich and interesting from a theoretical point of view and very important in real applications \cite{Marler2009,Hillermeier2005}. Multicriteria problems describe decision-making situations with multiple goals which are considered as functional criteria to be optimized.

@@

Finally, the last factor is the multiextremality of criteria \cite{Pardalos2017}. First of all, multiextremal criteria can generate complicated Pareto sets consisting, for example, of several disconnected parts. Moreover, the combination of such factors as multiextremality and greater dimension leads to significant computational complexity of the problem because the computational cost increases exponentially when dimension grows. From this point of view the multiextremal problems are the most complicated ones in multicriteria optimization and this MCO class is the subject of interest in the given paper.

@@

Another idea applies a parameterization scheme for the criteria of the problem and reduces a MCO problem to a family of single-criterion (scalar) optimization problems being problems of mathematical programming which can be solved by corresponding well-developed optimization methods \cite{Collette2004,Ehrgott2005}. In this scheme a non-negative coefficient (weight) is assigned to each criterion which reflects numerically an ``importance'' of the criterion and on the base of such weighted criteria a scalar function (convolution) is built and minimized. Under appropriate choice of convolution the solution of the built scalar optimization problem is a partial solution of the initial multicriteria problem. Changing weights we can obtain different Pareto points.

@@

The problem of multicriteria optimization is formulated as minimizing over the domain (\ref{Q}) the vector function
\begin{equation}\label{W}
W(x) = \left(w_1(x), ..., w_k(x)\right).
\end{equation}

@@

As mentioned in Introduction, one way of building the Pareto set consists in reducing the multiobjective problem to a family of single-criteria, or scalar optimization problems and solving them by methods of mathematical programming.  The global minimum point of a problem from this family will be a Pareto optimal point under corresponding assumptions. This approach can be realized by means of applying a convolution of the vector criterion (\ref{W}), for example, the maximum convolution
\begin{equation}\label{conv}
\Gamma(\lambda,x) = \max_{1 \leq i \leq k}{\lambda_i w_i(x)},
\end{equation}
where coefficients $\lambda_i, \; 1 \leq i \leq k,$ (weights of criteria) satisfy the conditions
\begin{equation}\label{lambda}
\lambda_i \geq 0, \; 1 \leq i \leq k, \; \sum_{i=1}^k{\lambda_i} = 1.
\end{equation}
If
\begin{equation}\label{positive}
w_i(x) > 0, \; 1 \leq i \leq k,
\end{equation}
in the domain (\ref{Q}), then global solution to the problem
\begin{equation}\label{conv_problem}
\Phi_\lambda(x) = \Gamma(\lambda,x) + \gamma \sum_{i=1}^k{\lambda_i w_i(x)} \rightarrow \min, \; x \in Q,
\end{equation}
with a sufficiently small parameter $\gamma > 0$ is a Pareto optimal point of the initial MCO problem (\ref{problem}) \cite{Wierzbicki,Marler2004}. The positiveness of criteria is not restrictive requirement because it is possible to transform easily the MCO problem with non-positive criteria to the form (\ref{positive}) without loss of Pareto solution.

@@

If the function $f(x)$ satisfies the Lipschitz condition (\ref{lip_1}) then $H=2L\sqrt{N+3}$.
As a consequence, minimization of the function $f(x(t))$  requires the use of special algorithms oriented at functions with property (\ref{holder}).

In the proposed scheme for solving reduced univariate minimization problem we apply modified information-statistical Algorithm \cite{Strongin2000} with Local Refinements (ALR) accelerating convergence to global optimum of functions that satisfy the H{\"o}lder condition.

@@

\item
The new $(n+1)$-th trial is performed at the point
\begin{equation}\label{eq:22}
t^{n+1} = \frac{t_p-t_{p-1}}{2} - sign(v_p-v_{p-1}) \left( \frac{|v_p-v_{p-1}|}{M} \right)^N \frac{1}{2r},
\end{equation}
the value $v^{n+1} = \varphi(t^{n+1})$ is computed and the iteration number $n$ is increased by 1.

@@

The method ALR described above is a modification of the basic information-statistical algorithm \cite{Strongin2000,Sergeyev2013} for minimization functions satisfying the H{\"o}lder condition. The modification consists in introducing new characteristics (\ref{eq:17}) after initial stage of the search to improve the method's speedup on account of increasing values of characteristics for intervals containing current record (\ref{eq:18}). 
Actually, the second term in (\ref{eq:17}) will have the largest value equal to $(1.5)^{\sigma}$ only for those intervals for which the equality $v_q = \varphi^*(n)$ or $v_{q-1} = \varphi^*(n)$ holds, i.e., for the intervals with the boundary points one of which coincides with the current record $\varphi^*(n)$.

@@

The complete convergence theory of the information-statistical algorithms is given in \cite{Strongin2000,Sergeyev2013}.

@@

During all experiments the method ALR was used with parameters $r=5.6$, $\sigma = 18$, $T=3$. 
The use of $T=3$ corresponds to alternation of one ``local'' iteration and two ``global'' iterations, that can be interpreted as a combination of local refinement and global updating of the current optimizer.
Accuracy $\varepsilon$ from the stopping condition (\ref{eq:23}) was $0.05$ in bi-criteria problems and $0.01$ in 4-criteria ones.

@@

The Pareto front in one of test problems is shown in Fig.~\ref{fig2a} and Fig.~\ref{fig2b}. Here Fig.~\ref{fig2a} corresponds to the Pareto front computed on the uniform grid with the step 0.05 in the search domain whereas Fig.~\ref{fig2b} shows the Pareto front computed by RALR-DA.

@@

The averaged HV indices obtained while solving the two-dimensional bi-criteria problems are presented in Tab.~\ref{tab1} and Fig.~\ref{fig3} where ``Pareto'' corresponds to the estimation obtained by the grid method.

@@

\begin{table}
\tbl{Results of testing 2-dimensional MCO problems with 4 criteria.}
{\begin{tabular}{ccccccc} \toprule
 Pareto & RALR-DA & OMOPSO & SPEA2 & SMPSO & IBEA & NSGAII   \\ \midrule
  23.89 & 23.52 & 22.87 & 22.80 & 22.53 & 20.68 & 19.82 \\ \bottomrule
\end{tabular}}
\label{tab3}
\end{table}

@@

As the directions of further development of the current research it would be interesting to investigate the possibilities of applying other schemes of dimensionality reduction, for example,
the nested optimization scheme \cite{Grishagin2016,Grishagin2018}, the diagonal \cite{Sergeyev2015,Sergeyev2017} or simplicial \cite{Zilinskas2011,PaulaviciusZilinskas2014} partition methods
in the framework of the general approach used in the paper. Moreover, the prospective way is to develop parallel versions of the proposed general scheme oriented at powerful supercomputer systems.
